


Problem A:
```{r}
library(numDeriv)

# Gradient Descent Function
gradientDescent <- function(f, starting_value, max_iter, gamma = 0.01) {
  current_value <- starting_value
  history <- numeric(max_iter) # To store function values for plotting
  
  for (i in 1:max_iter) {
    grad <- grad(f, current_value) # Compute the gradient at current value
    current_value <- current_value - gamma * grad # Update the current value
    history[i] <- f(current_value) # Store function value after update for plotting

    # Optional: Implement convergence criteria to break the loop early
    # if (norm(grad, type="2") < tolerance) {
    #    break
    # }
  }
  
  list(minimizer=current_value, history=history)
}

# Test functions
# Define the test function as per the formula given
test_function <- function(v) {
  x <- v[1]
  y <- v[2]
term1 <- (1 + (x + y + 1)^2 * (19 - 14*x + 3*x^2 - 14*y + 6*x*y + 3*y^2))
 term2 <- (30 + (2*x - 3*y)^2 * (18 - 32*x + 12*x^2 + 48*y - 36*x*y + 27*y^2))
  
  f_value <- term1 * term2
  return(f_value)
}

starting_value <- c(0, 0) # Starting at the origin
max_iter <- 10000 # Maximal number of iterations
gamma <- 0.0001 # Learning rate, might need adjustment based on the function characteristics

result <- gradientDescent(test_function, starting_value, max_iter, gamma)

print(paste("Approximate minimizer found at x =", result$minimizer[1], "y =", result$minimizer[2]))

# Plotting the path of function values
plot(result$history, type='l', main="Function Value Over Iterations", xlab="Iteration", ylab="Function Value",
     col="blue")
```


   
```


```{r}
starting_value <- c(0, 0) # Starting at the origin
max_iter <- 1000 # Maximal number of iterations
gamma <- 0.001 # Learning rate, might need adjustment based on the function characteristics

result <- gradientDescent(test_function, starting_value, max_iter, gamma)

print(paste("Approximate minimizer found at x =", result$minimizer[1], "y =", result$minimizer[2]))

# Plotting the path of function values
plot(result$history, type='l', main="Function Value Over Iterations", xlab="Iteration", ylab="Function Value",
     col="blue")




```

```{r}
test_function(c(0,-1))
```


